---
title: "Getting started with cofechar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cofechar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # examples require .rwl files not bundled with the package
)
```

## Overview

**cofechar** provides a faithful R port of two modules from Holmes'
Dendrochronology Program Library (DPL):

| Module | Functions | DPL routine |
|--------|-----------|-------------|
| EDT    | `dpl_read`, `dpl_write`, `dpl_read_dec`, `dpl_merge`, `dpl_edt`, `dpl_display`, `dpl_edit_file` | DPLEDT |
| COF    | `dpl_cof`, `dpl_cof_diag`, `dpl_dateme`, `dpl_barplot`, `plot_cof_barplot` | DPLCOF / QCOF |

All internal statistical routines — cubic smoothing spline, Burg AR modelling,
log-transform, normalisation — are translated directly from Holmes' Fortran
source (Holmes 1982–1994) so that numerical output is equivalent to the DPL
reference implementation.

---

## Typical workflow

### 1 — Read and inspect

```{r read}
library(cofechar)
library(dplR)

# Read the Cerro Mirador Fitzroya dataset (included in the package)
mir_file <- system.file("extdata", "CL-MIR.rwl", package = "cofechar")
rwl <- dpl_read_dec(mir_file, label_length = NULL, stop_val = -9999L,
                    unit = "0.001mm")

# Inspect a 50-year window around a suspect year
dpl_display(rwl, series = "ACC014A", around_yr = 1950)
```

### 2 — Edit

```{r edit}
rwl_ed <- dpl_edt(
  rwl,
  edits = list(
    # Replace a suspicious value in ACC014A (visible drought signal ~1953)
    list(series = "ACC014A", op = "replace", year = 1953, value = 0.26),

    # Trim the longest series to a common overlap start
    list(series = "ACC026B", op = "trim_start", first_year = 1856),
    list(series = "ACC026",  op = "trim_start", first_year = 1856),

    # Drop a low-correlation core after reviewing COFECHA diagnostics
    list(series = "ACC23TA", op = "omit")
  )
)
```

Available operations: `copy`, `omit`, `rename`, `first_year`, `last_year`,
`replace`, `insert`, `delete`, `trim_start`, `trim_end`.

### 3 — COFECHA quality control

```{r cof}
# Full run with default options — output mirrors DPL Parts 1–7
cof <- dpl_cof(rwl_ed, output_file = tempfile(fileext = ".out"))

# Inspect the statistics table (ranked by master correlation)
cof$stats[order(cof$stats$r_master), ]

# List all flagged segments
subset(cof$segments, flag != "")

# On-demand Part 6 diagnostics for the lowest-correlating series
worst <- cof$stats$series[which.min(cof$stats$r_master)]
dpl_cof_diag(cof, series = worst)
```

### 4 — Iterate

```{r iterate}
# After reviewing diagnostics, remove another problematic series and re-run
drop <- cof$stats$series[cof$stats$r_master < cof$crit]
rwl_ed2 <- dpl_edt(rwl_ed,
  edits   = lapply(drop, function(id) list(series = id, op = "omit")),
  verbose = FALSE)
cof2 <- dpl_cof(rwl_ed2, verbose = FALSE)
range(cof2$stats$r_master)
```

### 5 — Plot and export

```{r plot}
# ASCII bar plot of the COFECHA master (matches DPL Part 4)
dpl_barplot(cof)

# ASCII bar plot of a single series
dpl_barplot(rwl, series = "ACC026B", years = c(1850, 2002))

# Publication-quality graphical bar plot
pdf(tempfile(fileext = ".pdf"), width = 14, height = 4)
plot_cof_barplot(cof,
  main = "Fitzroya cupressoides — Cerro Mirador (Barichivich 2005)")
dev.off()
```

### 6 — Feed into dplR

All functions return standard dplR `rwl` data.frames:

```{r dplr}
crn <- dplR::chron(dplR::detrend(rwl_ed, method = "Spline"))
# rwl_ed is a standard dplR rwl — all dplR functions apply directly
```

---

## Dating undated series

```{r dateme}
dat <- dpl_dateme(rwl_undated, cof_result = cof)
dat$summary   # best_adj = year offset supported by most segments
```

---

## References

Holmes, R. L. (1983). Computer-assisted quality control in tree-ring dating
and measurement. *Tree-Ring Bulletin* 43:69–78.

Holmes, R. L. (1994). *Dendrochronology Program Library*. Laboratory of
Tree-Ring Research, University of Arizona, Tucson.
